---
title: "Functional Elsa"
output: html_document
date: "2023-10-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
## Load required libraries
library(ggnewscale) 
library(ggtree)
library(R.utils) 
library(tidyverse) 
library(ape)
library(devtools) 
library(ggplot2) 
library(hillR) 
library(spaa) 
library(vegan)
#library(hilldiv)
library(hilldiv2)
library(Rhdf5lib)
library(phyloseq)
library(phytools) 
library(microbiome) 
library(matrixStats)
library(microbiomeutilities) 
library(lme4) 
library(MuMIn) 
library(nlme) 
library(knitr) 
library(kableExtra)
library(pairwiseAdonis)
library(sjPlot)
library(distillR) 
library(RColorBrewer) 
library(reshape2) 
library(ggpubr) 
library(ggdendro) 
library(grid)
library(gplots)
library(dendextend)
library(stringr) 
library(Rtsne)
library(glue)
library(webshot2)
library(dplyr)
library(POMS)
```


```{r directories, comment="", echo=FALSE, message=FALSE, warning=FALSE}
## Declare directories and files

workingdir="/Users/elsa/Desktop/THESIS/Thesis"
counts_file="/Users/elsa/Desktop/THESIS/Thesis/DMB0032_counts.tsv"
tree_file="/Users/elsa/Desktop/THESIS/Thesis/DMB0032.tree"
taxonomy_file="/Users/elsa/Desktop/THESIS/Thesis/DMB0032_mag_info.tsv"
metadata_file="/Users/elsa/Desktop/THESIS/Thesis/DMB0032_metadata.tsv"
coverage_file="/Users/elsa/Desktop/THESIS/Thesis/DMB0032_coverage.tsv"
```

```{r loaddata, comment="", echo=FALSE, message=FALSE, warning=FALSE}
## Load data 
setwd(workingdir) 
counts_table <- read.table(counts_file, sep="\t",row.names=1,header=T)
coverage_table <- read.table(coverage_file, sep="\t",row.names=1,header=T)
metadata <- read.table(metadata_file,sep="\t",header=T)%>%
	rename(sample=EHI_plaintext)
mags_table <- read.table(taxonomy_file,sep="\t",header=T)
tree <- read.tree(tree_file)

ehi_phylum_colors <- read.table("~/Desktop/Thesis/R-analysis1/ehi_phylum_colors.tsv",sep="\t",header=T,comment.char = "")
#Delete *__ from the taxonomy names
ehi_phylum_colors1 <- ehi_phylum_colors %>%
  mutate_at(vars(phylum), ~ str_replace(., "[dpcofgs]__", ""))

taxonomyclean <- mags_table%>%
  mutate_at(vars(domain,phylum,class,order,family,genus,species), ~ str_replace(., "[dpcofgs]__", ""))

metadata_unified <- read.csv("/Users/elsa/Desktop/THESIS/Thesis/metadata_unified.csv")
metadata_unified_clean <- metadata_unified %>%
  filter(sample != "")  # Filter rows where "sample" is not an empty string
selected_columns <- metadata_unified_clean %>%
  select(sample, Pack, PR_results_short, toxocaris_leonina, sarcocystis_spp, Tania_serialis, cryptosporidium, giardia, toxoplasma_gondii, trichincella_spp, Huldscore)
# Merge the selected columns with "metadata" based on the "sample" column
all_metadata <- left_join(metadata, selected_columns, by = "sample")

min_coverage=0.3
count_table_cov <- coverage_table %>%
  mutate(across(everything(), ~ ifelse(. > min_coverage, 1, 0))) %>%
  map2_df(., counts_table, ~ .x * .y) %>%
  as.data.frame()
rownames(count_table_cov) <- rownames(coverage_table)

mags_table1 <- column_to_rownames(mags_table, "genome")
genome_read_sizes <- mags_table1[rownames(count_table_cov),] %>%
    select(mag_size) %>%
    mutate(mag_size = mag_size / 143) %>% #143 nt is the average read-length after quality filtering in EHI data
    pull()
count_table_cov_size <- sweep(count_table_cov, 1, genome_read_sizes, "/")

count_table_cov_size_rel <- count_table_cov_size %>%
  rownames_to_column("Genome") %>%
  mutate_at(vars(-Genome),~./sum(.)) #TSS normalisation

count_filtered <- column_to_rownames(count_table_cov_size_rel, "Genome")
```

## Functional analysis

```{r filter_annotations, comment="", echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
#### read table ####
filtered_df <- read.table("all_annotations.csv", sep="")
```


```{r gifts, comment="", echo=TRUE, message=FALSE, warning=FALSE, results='hide', cache=TRUE}
count_filtered_rel <- count_filtered %>%
  rownames_to_column(., "Genome") %>%
  mutate_at(vars(-Genome),~./sum(.))  %>%
  column_to_rownames(., "Genome")

#Run distillation
GIFTs <- distill(filtered_df,GIFT_db,genomecol=2,annotcol=c(9,10,19))
#GIFTs <- GIFTs[-c(9,10),]

# GIFTs_filtered <- GIFTs[rownames(GIFTs) %in% rownames(count_filtered_rel),]

#Aggregate bundle-level GIFTs into the compound level
GIFTs_elements <- to.elements(GIFTs,GIFT_db)

#Aggregate element-level GIFTs into the function level
GIFTs_functions <- to.functions(GIFTs_elements,GIFT_db)

#Aggregate function-level GIFTs into overall Biosynthesis, Degradation and Structural GIFTs
GIFTs_domains <- to.domains(GIFTs_functions,GIFT_db)

#Get community-weighed average GIFTs per sample
GIFTs_elements_community <- to.community(GIFTs_elements,count_filtered_rel,GIFT_db)
GIFTs_functions_community <- to.community(GIFTs_functions,count_filtered_rel,GIFT_db)
GIFTs_domains_community <- to.community(GIFTs_domains,count_filtered_rel,GIFT_db)
```

```{r alpha_div_funct, comment="", echo=TRUE, message=FALSE, warning=FALSE, results='hide'}

not_shared_rows <- setdiff(rownames(count_table_cov_size), rownames(GIFTs_elements)) #"EHA01909_bin.13" "EHA01928_bin.9" 
rows_to_remove <- c("EHA01909_bin.13", "EHA01928_bin.9")
count_table_cov_size_filtered <- count_table_cov_size[!rownames(count_table_cov_size) %in% rows_to_remove, ]
# There are 2 bins that are present in count_filtered_rel (555) but not GIFTs elements (553). I removed them for analysis, but I am questioning why they are not there, and if removing them is valid for analysis. 

#Convert traits into distance matrix
dist <- traits2dist(GIFTs_elements, method="gower")

functional_div_A <- hilldiv(data=count_filtered_rel,dist=dist)

#Q (Rao's Q); D_q (functional hill number, the effective number of equally abundant and functionally equally distinct species); MD_q (mean functional diversity per species, the effective sum of pairwise distances between a fixed species and all other species); FD_q (total functional diversity, the effective total functional distance between species of the assemblage).
```
ELSA QUESTION: There are 2 bins that are present in count_filtered_rel (555) but not GIFTs elements (553). I removed them for analysis, but I am questioning why they are not there, and if removing them is valid for analysis

#### Average functional diversities

```{r div_F_mean, comment="", echo=TRUE, message=FALSE, warning=FALSE}
alpha_div_func <-  t(functional_div_A) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("sample") %>%
  merge(., all_metadata, by="sample")

table_mean_alpha_func <- alpha_div_func %>%
  group_by(region) %>% 
  summarise_at(.vars = names(.)[2],.funs = c(mean="mean", sd="sd"))

knitr::kable(table_mean_alpha_func, format = "html", full_width = F,col.names = c('Groups', 'Mean', 'SD'), digits = 3) %>%
  kable_styling(latex_options="scale_down")

```

#### Functional diversity plots

```{r alpha_div_func_plot1, comment="", echo=TRUE, message=FALSE, warning=FALSE}
alpha_div_func %>%
  ggplot(aes(x = region, y = q1, group = region, color = region)) +
  geom_boxplot(alpha = 0.2, outlier.shape = NA, width = 0.3, show.legend = FALSE, coef = 0) +
  geom_jitter(width = 0.1, show.legend = TRUE) +
  theme(panel.background = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(size = 0.5, linetype = "solid", colour = "black")) +
  xlab("Region") + 
  ylab("Functional diversity")
```

```{r}
# Shapiro-Wilk test for normality
alpha_div_func %>%
  filter(region %in% c("Daneborg", "Ittoqqortoormii")) %>%
  group_by(region) %>%
  summarise(shapiro_p = shapiro.test(q1)$p.value) -> shapiro_results_F

# Results: p =  7.498381e-09	(Daneborg), p = 2.627117e-03	 Ittoqqortoormiit
# THEREFORE the data is NOT NORMALLY distributed

# Using leveneTest()
car:: leveneTest(q1 ~ region, alpha_div_func)
# Results: p-value = 0.03988
# THEREFORE therefore equal variance cannot be assumed

# Wilcoxon signed rank test for statistical significance
alpha_div_func %>%
  filter(region %in% c("Daneborg", "Ittoqqortoormii")) %>%
  group_by(region) %>%
  summarise(wilcox_p = wilcox.test(q1 ~ region, data = .)$p.value) -> wilcox_results_F

print(wilcox_results_F)
# Results: p = 0.0001328166	
# THEREFORE the difference in alpha diversity between Ittoq and Daneborg IS considered statistically significant. 
```


## Functional beta diversity calculations
```{r beta_div_func, comment="", echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
beta_q1f <- hilldiv2::hillpair(count_table_cov_size_filtered, dist=dist) %>%
  list2dist()
```

```{r}
sample_table_adonis_func <- all_metadata %>%
    filter(sample %in% labels(beta_q1f)) %>%
    arrange(sample) %>%
    # mutate(location=paste0(round(longitude,2),"_",round(latitude,2))) %>%
    # select(sample,region) %>%
    select_if(~ length(unique(.)) > 1) %>% #remove columns with all-identical values
    column_to_rownames(var = "sample") %>%
    as.data.frame()
```
#### NMDS:
```{r}
beta_q1f_nmds <- beta_q1f %>%
                metaMDS(.,trymax = 500, k=2, verbosity=FALSE) %>%
                vegan::scores() %>%
                as_tibble(., rownames = "sample") %>%
                left_join(all_metadata, by = join_by(sample == sample))
```
```{r}
group_f <- length(unique(beta_q1f_nmds$region))

beta_q1f_nmds %>%
            group_by(region) %>%
            mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
            mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
            ungroup() %>%
            ggplot(., aes(x=NMDS1,y=NMDS2, color=region)) +
                scale_color_manual(values=beta_colors[c(1:group_n)]) +
                geom_point(size=2) +
                geom_segment(aes(x=x_cen, y=y_cen, xend=NMDS1, yend=NMDS2), alpha=0.2) +
                theme_classic() +
                theme(legend.position="right", legend.box="vertical") +
                guides(color=guide_legend(title="Region"))
```

Adonis
```{r Adonis_Region, comment="", echo=TRUE, message=FALSE, warning=FALSE}
# order data
m3 <- as.matrix(beta_q1f) 
sample_table_adonis_row_func <- rownames_to_column(sample_table_adonis_func, "Samples")
metadata_ord_3<- sample_table_adonis_func[order(match(sample_table_adonis_row_func$Sample,rownames(m3))),]

#Permutation test
ps.disper_func <- betadisper(beta_q1f, metadata_ord_3$region) 
permutest(ps.disper.neutral, pairwise = TRUE) 
# F Statistic: 
# p-value: 

# Adonis
adonis2(beta_q1f ~ region, data =metadata_ord_3, permutations = 999)
# A p-value of <0.05 is significant. The R2 value will tell you what amount of variation is based off the clustering variable. in this case the region. 
# p-value:
# R2: 
```
## Functional capacity of the MAGs

```{r GIFTs_elements, comment="", echo=FALSE, message=FALSE, warning=FALSE, fig.dim = c(20, 20)}
GIFTs_elements %>%
  reshape2::melt() %>%
  rename(Genome = Var1, Code_element = Var2, GIFT = value) %>%
  inner_join(GIFT_db,by="Code_element") %>%
  ggplot(., aes(x=Code_element, y=Genome, fill=GIFT, group=Code_function))+
  geom_tile()+
  scale_y_discrete(guide = guide_axis(check.overlap = TRUE))+
  scale_x_discrete(guide = guide_axis(check.overlap = TRUE))+
  scale_fill_gradientn(limits = c(0,1), colours=brewer.pal(7, "YlGnBu"))+
  facet_grid(. ~ Code_function, scales = "free", space = "free")+
  theme_grey(base_size=8)+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),strip.text.x = element_text(angle = 90))
```

```{r hierachical_clustering, comment="", echo=FALSE, message=FALSE, warning=FALSE, fig.dim = c(20, 20)}
col <- colorRampPalette(brewer.pal(7, "YlGnBu"))(256)
heatmap(GIFTs_functions_community, scale = "none", col=col, Colv=NA,
        trace = "none", density.info = "none", ylab="Samples", 
        xlab="Gifts", cexRow=0.7, margins = c(8,8), lhei=c(2,4), lwid=c(2,5), 
        keysize=0.75, key.par = list(cex=0.9),srtCol = 45)
```


```{r tsne_prep, comment="", echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
# Split name column into first name and last name
taxonomy <- taxonomyclean # %>%
#  select(user_genome, classification) %>%
#  rename(Genome = user_genome) %>%
#  separate(classification, c('Domain', 'Phylum','Class','Order','Family','Genus','Species'),sep=";") %>%
#  mutate_at(vars(Domain,Phylum,Class,Order,Family,Genus,Species), ~ str_replace(., "[dpcofgs]__", "")) %>%
#  mutate_at(vars(Genome), ~ str_replace(., ".fa", ""))
```

```{r tsne, comment="", echo=FALSE, message=FALSE, warning=FALSE}
set.seed(100)

GIFTs_elements_tSNE <- Rtsne(X=GIFTs_elements, dims = 2, perplexity = floor((nrow(GIFTs_elements) - 1) / 3), check_duplicates = FALSE)

GIFTs_domains <- as.data.frame(GIFTs_domains)
GIFTs_domains$genome <- rownames(GIFTs_domains)

GIFTs_elements_tSNE <- GIFTs_elements_tSNE$Y %>%
  as.data.frame() %>%
  mutate(genome=rownames(GIFTs_elements)) %>%
  inner_join(GIFTs_domains, by="genome") %>%
  inner_join(taxonomyclean, by="genome") %>%
  mutate_at(vars(phylum, class, order, family, genus), factor) %>%
  #mutate(cyl = factor(Phylum, levels = phylum_colors$Phylum)) %>%
  rename(tSNE1="V1", tSNE2="V2")

count_t <- count_filtered_rel %>%
  t() %>%
  as.data.frame()

#colnames(count_t) <- count_t[1,]
#count_t <- count_t[-1, ]

```

Create Merge table and GIFT elements for Ittoq dogs
```{r tsne_Ittoq, comment="", echo=FALSE, message=FALSE, warning=FALSE}
mergetable_Ittoq <- count_t %>%
  mutate_if(is.character, as.numeric) %>%
  rownames_to_column("sample") %>%
  merge(., all_metadata[ , c("sample", "region")], by="sample", all.x = TRUE) %>%
  filter(region=="Ittoqqortoormii") %>%
  column_to_rownames(., "sample")

mergetable_Ittoq <- mergetable_Ittoq [1: ncol(mergetable_Ittoq)-1 ]

GIFTs_elements_tSNE_rel_means_I <- colMeans(mergetable_Ittoq) %>%
  as.data.frame() %>%
  rename( "Relative_value" = ".")  %>%
  rownames_to_column(., "genome") %>%
  merge(., GIFTs_elements_tSNE, by= "genome")

GIFTs_elements_tSNE_rel_means_Ittoq <- GIFTs_elements_tSNE_rel_means_I[GIFTs_elements_tSNE_rel_means_I$Relative_value != 0, ]

```

Create Merge table and GIFT elements for Daneborg dogs
```{r tsne_Daneborg, comment="", echo=FALSE, message=FALSE, warning=FALSE}
mergetable_Daneborg <- count_t %>%
  mutate_if(is.character, as.numeric) %>%
  rownames_to_column("sample") %>%
  merge(., all_metadata[ , c("sample", "region")], by="sample", all.x = TRUE) %>%
  filter(region=="Daneborg") %>%
  column_to_rownames(., "sample")

mergetable_Daneborg <- mergetable_Daneborg [1: ncol(mergetable_Daneborg)-1 ]

GIFTs_elements_tSNE_rel_means_D <- colMeans(mergetable_Daneborg) %>%
  as.data.frame() %>%
  rename( "Relative_value" = ".")  %>%
  rownames_to_column(., "genome") %>%
  merge(., GIFTs_elements_tSNE, by= "genome")

GIFTs_elements_tSNE_rel_means_Daneborg <- GIFTs_elements_tSNE_rel_means_D[GIFTs_elements_tSNE_rel_means_D$Relative_value != 0, ]
```

### Region Daneborg: tSNE Phylum

```{r tsneIttoq_plot_phylum, comment="", echo=FALSE, message=FALSE, warning=FALSE}
GIFTs_elements_tSNE_rel_means_Ittoq %>%
  ggplot(aes(x = tSNE1, y = tSNE2, color = phylum))+
  geom_point(aes(size = Relative_value), shape=16, alpha=0.8) +
#  scale_color_gradientn(limits = c(0,1), colours=brewer.pal(7, "YlGnBu"))+
  theme_minimal() +
  theme()
```

### Region Daneborg: tSNE Phylum

```{r tsneDaneborg_plot_phylum, comment="", echo=FALSE, message=FALSE, warning=FALSE}
GIFTs_elements_tSNE_rel_means_Daneborg %>%
  ggplot(aes(x = tSNE1, y = tSNE2, color = phylum))+
  geom_point(aes(size = Relative_value), shape=16, alpha=0.8) +
  #scale_color_gradientn(limits = c(0,1), colours=brewer.pal(7, "YlGnBu"))+
  theme_minimal() +
  theme()
```

### Region Ittoq: tSNE gifts

```{r tsneIttoq_plot_gift, comment="", echo=FALSE, message=FALSE, warning=FALSE}
GIFTs_elements_tSNE_rel_means_Ittoq %>%
  ggplot(aes(x = tSNE1, y = tSNE2, color = Overall))+ #coloured by GIFT
  geom_point(aes(size = Relative_value), shape=16, alpha=0.8) +
  scale_color_gradientn(limits = c(0,1), colours=brewer.pal(7, "YlGnBu"))+
  theme_minimal() +
  theme()
```

### Region Daneborg: tSNE gifts

```{r tsneDaneborg_plot_gift, comment="", echo=FALSE, message=FALSE, warning=FALSE}
GIFTs_elements_tSNE_rel_means_Daneborg %>%
  ggplot(aes(x = tSNE1, y = tSNE2, color = Overall))+ 
  geom_point(aes(size = Relative_value), shape=16, alpha=0.8) +
  scale_color_gradientn(limits = c(0,1), colours=brewer.pal(7, "YlGnBu"))+
  theme_minimal() +
  theme()
```


POMS:
# AS FROM https://academic.oup.com/bioinformatics/article/38/22/5055/6731923?login=true (on the stats discord)
```{r}
taxa_abun <- count_filtered_rel
func <- GIFTs_elements_community %>%
  as.data.frame() %>%
  t()
func <- as.data.frame(func)
tree <- as.phylo(tree) # something is wrong with the structure of the tree, otherwise the rest should run
Ittoq_samples <- all_metadata %>%
  filter(region == "Ittoqqortoormii") %>%
  pull(sample)
Daneborg_samples <- all_metadata %>%
  filter(region == "Daneborg") %>%
  pull(sample)

# Example of how to run main POMS function. 
POMS_out <- POMS_pipeline(abun = taxa_abun,
                          func = func,
                          tree = tree,
                          group1_samples = Ittoq_samples,
                          group2_samples = Daneborg_samples,
                          ncores = 1,
                          pseudocount = 0.000001,
                          verbose = TRUE)
```

